Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-29 paper.tex
Put back ofh website ref

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data, encoding information about
DNA sequence polymorphisms among a set of samples with associated
quality control metrics and metadata~\citep{danecek2011variant}.
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
Our Future Health~\citep{cook2025our},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for millions of humans~\citep{stark2024call}.
Large genetic variation datasets are also being generated for other organisms
and a variety of purposes including
agriculture~\citep{ros2020accuracy,wang2023rice},
conservation~\citep{shaffer2022landscape}
and infectious disease surveillance~\citep{hamid2023pf7}.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} make it an
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second}
or BGEN~\citep{band2018bgen}
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use bespoke binary formats
that support the required access patterns~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one,
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate}.
While PLINK and BGEN formats are more efficient to access than VCF, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype
data, but file sizes are substantially larger than
the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
For example, at two bits
per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

@@

\subsection{Case study: Our Future Health genotype data}
Our Future Health~\cite{cook2025our}  is the UK’s largest ever health research programme.
It is designed to help people live healthier lives for longer through the
discovery and testing of more effective approaches to prevention, earlier
detection and treatment of diseases.
Data Release 9~\citep{ofhdatarelease9} was made available in December 2024,
including genotype array data for from 651,050 individuals
and 707,522 variants.
For this study, we converted the genotype data for chromosome 22 consisting 
of 10,221 variants over three gzipped VCF files and a total of
42.3 GiB of storage. 
% 20G     ofh.chr22-b0.vcf.gz
% 19G     ofh.chr22-b1.vcf.gz
% 3.3G    ofh.chr22-b2.vcf.gz
% ==42.3 GiB - assuming the G is power-of-two here
% 10,221 variants and 42.3 GiB of storage.

>>>>

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data, encoding information about
DNA sequence polymorphisms among a set of samples with associated
quality control metrics and metadata~\citep{danecek2011variant}.
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
Our Future Health~\citep{cook2025our,ofhmain_2025},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for millions of humans~\citep{stark2024call}.
Large genetic variation datasets are also being generated for other organisms
and a variety of purposes including
agriculture~\citep{ros2020accuracy,wang2023rice},
conservation~\citep{shaffer2022landscape}
and infectious disease surveillance~\citep{hamid2023pf7}.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} make it an
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second}
or BGEN~\citep{band2018bgen}
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use bespoke binary formats
that support the required access patterns~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one,
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate}.
While PLINK and BGEN formats are more efficient to access than VCF, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype
data, but file sizes are substantially larger than
the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
For example, at two bits
per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

@@

\subsection{Case study: Our Future Health genotype data}
Our Future Health~\cite{cook2025our,ofhmain_2025}  is the UK’s largest ever health research programme.
It is designed to help people live healthier lives for longer through the
discovery and testing of more effective approaches to prevention, earlier
detection and treatment of diseases.
Data Release 9~\citep{ofhdatarelease9} was made available in December 2024,
including genotype array data for from 651,050 individuals
and 707,522 variants.
For this study, we converted the genotype data for chromosome 22 consisting 
of 10,221 variants over three gzipped VCF files and a total of
42.3 GiB of storage. 
% 20G     ofh.chr22-b0.vcf.gz
% 19G     ofh.chr22-b1.vcf.gz
% 3.3G    ofh.chr22-b2.vcf.gz
% ==42.3 GiB - assuming the G is power-of-two here
% 10,221 variants and 42.3 GiB of storage.
